---
title: "Final Analysis before moving to markdown"
output:
  html_document:
    df_print: paged
---


```{r}
library(tidyverse)
library(tsibbledata)
library(tsibble)
library(lubridate)
library(ggplot2)
library(ggthemes)
```


```{r}
nyc_bikes_df <- tsibbledata::nyc_bikes
nyc_bikes_df
# glimpse(nyc_bikes_df)
# view(nyc_bikes_df)
```

1. Adding columns: start day of the week, month, time
2. Creating a column that shows the season
3. Creating a column for age of rider and removing two outliers (born 1880s)
4. Showing hire durations

```{r}
#creating 4 columns: date, and start day, month and hour
nyc_bikes_df <- nyc_bikes_df %>% 
  mutate(date = as_date(start_time), .after = 2,
         start_day = wday(start_time, label = TRUE, abbr = FALSE),
         start_month = month(start_time, label = TRUE, abbr = FALSE),
         start_hour = hour(start_time), minute(start_time))
#Couldn't unite the start time hours and mins but I am only going to look at 
#hour anyway so it doesn't matter:
nyc_bikes_df <- nyc_bikes_df %>% 
  unite(start_hire_time, 
        start_hour, `minute(start_time)`, 
        sep = ":", 
        remove = FALSE)
nyc_bikes_df

#adding season column just in case
nyc_bikes_df <- nyc_bikes_df %>% 
  mutate(season = case_when(
    start_month %in% c("December", "January", "February") ~ "Winter",
    start_month %in% c("March", "April", "May") ~ "Spring",
    start_month %in% c("June", "July", "August") ~ "Summer",
    start_month %in% c("September", "October", "November") ~ "Autumn",
    ), .after = 5)
nyc_bikes_df

# showing duration of hire (2 were hired for 3 days and the rest less than a day
# most were hired for less than 50 mins)
nyc_bikes_df <- nyc_bikes_df %>% 
  mutate(hire_duration = stop_time - start_time, # this shows 1.50(= 1min 30sec)
         length_of_hire_rounded = as.period(round(stop_time - start_time)), 
         hours_hired = as.numeric(length_of_hire_rounded, "hours")
  )
nyc_bikes_df

#Finding people's ages in 2018 - 2018 was the year the data was collected
nyc_bikes_df <- nyc_bikes_df %>% 
  mutate(age_years = as.numeric(2018 - birth_year))  
nyc_bikes_df

# There are two rows where the customer is born in the 1880s. That can not be
# right so I am removing those two rows (neither has a gender either)
nyc_bikes_df <- nyc_bikes_df %>% 
  filter(birth_year > 1900)
nyc_bikes_df


```

Looking at subscribers, customers and unknown genders
```{r}
# 3951 are subscribers 
nyc_bikes_df %>% 
  filter(type == "Subscriber")

# 315 are customers
nyc_bikes_df %>% 
  filter(type == "Customer")

# 267 have gender = "unknown" and I noticed that most are 49 years old (n = 243)
nyc_bikes_df %>% 
  filter(gender == "Unknown" & age_years == 49)

# 49 year olds:
# 243/269 have gender unknown (26 have a gender).
# 89/269 are subscribers, the rest (180) are customers (out of 315 in total)
age49 <- nyc_bikes_df %>% 
  filter(age_years == 49)
# view(age49)

```


### What is the pattern of bike hires over time (e.g. within a year, month, 
week, or day)?

```{r}
# view(nyc_bikes_df)
# 4,266 rows
```

Across the year, total bikes hired = 4,266 rows 


Total rentals per day:
```{r}

count_per_day <- nyc_bikes_df %>% 
  group_by(date) %>% 
  summarise(rentals_per_day = n()) %>% 
  arrange(desc(rentals_per_day))
count_per_day


count_per_day %>% 
  ggplot()+
  aes(x = date, y = rentals_per_day)+
  geom_line() + 
  labs(x = "\nDays in 2018",
       y = "Total number of rentals per day",
       title = "Total number of bike rentals per day",
       subtitle = "For every day in the year 2018")+
  theme_bw()

```

From the graph, we can see an increase in total rentals per day as we head 
from winter into summer (the spike is on 31st July, a Tuesday) and then a 
decrease again as temperatures cool as we head into autumn and winter again.  

Though a trend can be seen, it is a bit messy. To look at it another way we 
can look at the total bikes rented per month or by season:
```{r}
count_per_month <- nyc_bikes_df %>% 
  group_by(season, start_month) %>% 
  summarise(rentals_per_month = n()) %>% 
  arrange(desc(rentals_per_month))
count_per_month


count_per_month %>% 
  ggplot()+
  aes(x = start_month, y = rentals_per_month) +
  geom_col(fill = "#007cb9")+
  theme(axis.text.x = element_text(angle=45, hjust=0.9))+
  geom_text(aes(label = rentals_per_month),
    position = position_dodge(0.9),
    vjust = 0)+ 
  labs(x = "Month (in 2018)", 
       y = "Total rentals per month", 
       title = "Total Bike Rentals per Month in 2018")

count_per_month %>% 
  ggplot()+
  aes(x = season, y = rentals_per_month)+
  geom_col(fill = "#007cb9") + 
  labs(x = "Month (in 2018)", 
       y = "Total rentals per month", 
       title = "Total Bike Rentals per Month in 2018")
```


Total bikes hired per day:
```{r}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

#Tues: 720!, Fri 658, Wed, Thurs, Mon 628, Sat 523, Sun(453!))
bikes_per_day <- nyc_bikes_df %>% 
  group_by(start_day) %>% 
  summarise(num_of_bikes_hired = n()) %>% 
  arrange(desc(num_of_bikes_hired))
bikes_per_day

bikes_per_day %>%  
  ggplot(aes(x = start_day, y = num_of_bikes_hired))+
  scale_x_discrete(limits = weekdays) +
    geom_col()+
  theme(axis.text.x = element_text(angle=45, hjust=0.9))+
  geom_text(aes(label = num_of_bikes_hired),
    position = position_dodge(0.9),
    vjust = 0)


```

From above, we can see that weekdays are more popular bike hire days than 
weekends. 720 bikes were hired on Tuesdays compared with 453 on Sundays. Was 
this the case every month? 

```{r}
monthly_bikes_per_day <-nyc_bikes_df %>% 
  group_by(season, start_month, start_day) %>% 
  summarise(num_of_bikes_hired = n())
monthly_bikes_per_day


monthly_bikes_per_day %>% 
  ggplot()+
  aes(x = start_day, y = num_of_bikes_hired, 
      fill = start_day)+
  scale_x_discrete(limits = weekdays) +
  geom_col() + 
  facet_wrap(~start_month)+
  theme(axis.text.x = element_text(angle=45, hjust=0.9))

monthly_bikes_per_day %>% 
  ggplot()+
  aes(x = start_day, y = num_of_bikes_hired, 
      group = start_month, colour = start_month)+
  scale_x_discrete(limits = weekdays) +
  geom_line()+
  facet_wrap(~season)+
  theme(axis.text.x = element_text(angle=45, hjust=0.9))
```



Are people mostly travelling at commuting times then? Looking at the full year:
```{r}
time_of_ride <- nyc_bikes_df %>% 
  group_by(start_hour) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))
time_of_ride

time_of_ride %>% 
  ggplot()+
  aes(x = start_hour, y = count)+
  geom_col(fill = "#007cb9")+
  theme_bw()+
  geom_text(aes(label = count),
    position = position_dodge(0.9),
    vjust = 0)+
  scale_x_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24)) + 
  labs(x = "\nHire Start Time (24hr clock", 
       y = "Total number of hires", 
       title = "Total number of hires by hour across the year 2018")

```
Looking by day and month: 
```{r}
#month
time_of_ride_by_month <- nyc_bikes_df %>% 
  group_by(start_month, start_hour) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))
time_of_ride

time_of_ride_by_month %>% 
  ggplot()+
  aes(x = start_hour, y = count)+
  geom_col()+
  facet_wrap(~start_month)
```


```{r}
#day
time_of_ride_by_day <- nyc_bikes_df %>% 
  group_by(start_day, start_hour) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))
time_of_ride

time_of_ride_by_day %>% 
  ggplot()+
  aes(x = start_hour, y = count)+
  geom_col()+
  facet_wrap(~start_day)
```




___
## Demographics 

```{r}
#there are 3069 males, 930 females, 267 unknown (243 of them are "49" years old)
nyc_bikes_df %>% 
  group_by(gender) %>% 
  summarise(count = n())

# 3,999 rows when taking out gender unknown
minus_gender_unknown <- nyc_bikes_df %>% 
  filter(gender %in% c("Male", "Female"))
minus_gender_unknown

```

How many trips per gender?
```{r}

rentals_by_month_and_gender <- minus_gender_unknown %>% 
  group_by(start_month, gender) %>% 
  summarise(rentals_per_month = n()) 
rentals_by_month_and_gender

rentals_by_month_and_gender %>% 
  ggplot()+
  aes(x = start_month, y = rentals_per_month, group = gender, colour = gender)+
  geom_point()+
  geom_line() +
  theme(axis.text.x = element_text(angle=45, hjust=0.9))+
  geom_text(aes(label = rentals_per_month),
    position = position_dodge(1),
    vjust = 0)
  
# rentals_by_gender <- minus_gender_unknown %>% 
#   group_by(date, start_month, gender) %>% 
#   summarise(rentals_per_day = n()) 
# rentals_by_gender
# 
# rentals_by_gender %>% 
#   ggplot()+
#   aes(x = date, y = rentals_per_day)+
#   geom_point() +
#   facet_grid(~gender)+
#   theme(axis.text.x = element_text(angle=45, hjust=0.9))
# 
# rentals_by_gender %>% 
#   filter(gender == "Male") %>% 
#   ggplot()+
#   aes(x = date, y = rentals_per_day)+
#   geom_point()
# 
#   rentals_by_gender %>% 
#   filter(gender == "Female") %>% 
#   ggplot()+
#   aes(x = date, y = rentals_per_day)+
#   geom_point()+
#   theme(axis.text.x = element_text(angle=45, hjust=0.9))
```
February 2018 was the warmest on record... 



Do males and females hire at different times of year/ days of week ? 


```{r}
#Tues: 720!, Fri 658, Wed, Thurs, Mon 628, Sat 523, Sun(453!))
day_by_gender <- minus_gender_unknown %>% 
  group_by(start_day, gender) %>% 
  summarise(daily_users = n())
day_by_gender

day_by_gender %>% 
  ggplot(aes(x = start_day, y = daily_users, fill = gender))+
  scale_x_discrete(limits = weekdays) +
  geom_col(position = "dodge")+
  theme(axis.text.x = element_text(angle=45, hjust=0.9))+
  geom_text(aes(label = daily_users),
    position = position_dodge(0.9),
    vjust = 0)
```

```{r}
month_by_gender <- minus_gender_unknown %>% 
  group_by(start_month, start_day, gender) %>% 
  summarise(daily_users = n())
month_by_gender


month_by_gender %>% 
  ggplot(aes(x = start_day, y = daily_users, fill = gender))+
  scale_x_discrete(limits = weekdays) +
  geom_col(position = "dodge") +
  facet_wrap(~start_month)+
  theme(axis.text.x = element_text(angle=45, hjust=0.9))
```

Difference in gender start location? (stereotype for business vs holiday)
```{r}
# same top 2 anyway 
minus_gender_unknown %>% 
  group_by(start_station, gender) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))

```


### AGE 

How many rides per age
```{r}
nyc_bikes_df %>% 
  group_by(age_years) %>% 
  ggplot()+
  aes(x = age_years) +
  geom_bar()
```

rides per age by gender
```{r}
minus_gender_unknown %>% 
  group_by(age_years, gender) %>% 
  ggplot()+
  aes(x = age_years, fill = gender) +
  geom_bar()
```


## Type of trip

subscriber or not:
```{r}
# nyc_bikes_df %>% 
#   filter(type == "Subscriber")
# 
# nyc_bikes_df %>% 
#   filter(type == "Customer")
# 
# nyc_bikes_df %>% 
#   filter(is.na(type))
# 3953 rows are subscribers (true) 
# and 315 are customers (FALSE)
# 0 NAs.
# with the 49 year olds: 89/269 are subscribers, the rest (180) are customers (out of 315 in total)
```


DURATION:
```{r}
duration <- nyc_bikes_df %>% 
  group_by(length_of_hire_rounded) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count)) 

a <- nyc_bikes_df %>% 
  mutate(duration_block = case_when(
    length_of_hire_rounded <= "5M 0S" ~ "0 - 5", 
    length_of_hire_rounded <= "10M 0S" & 
      length_of_hire_rounded > "5M 0S" ~ "5 - 10",
    length_of_hire_rounded <= "15M 0S" & 
      length_of_hire_rounded > "10M 0S" ~ "10 - 15",
    length_of_hire_rounded <= "20M 0S" & 
      length_of_hire_rounded > "15M 0S" ~ "15 - 20",
    length_of_hire_rounded <= "25M 0S" & 
      length_of_hire_rounded > "20M 0S" ~ "20 - 25",
    length_of_hire_rounded <= "30M 0S" & 
      length_of_hire_rounded > "25M 0S" ~ "25 - 30",
    length_of_hire_rounded <= "35M 0S" & 
      length_of_hire_rounded > "30M 0S" ~ "30 - 35",
    length_of_hire_rounded <= "40M 0S" & 
      length_of_hire_rounded > "35M 0S" ~ "35 - 40",
    length_of_hire_rounded <= "45M 0S" & 
      length_of_hire_rounded > "40M 0S" ~ "40 - 45",
    length_of_hire_rounded <= "50M 0S" & 
      length_of_hire_rounded > "45M 0S" ~ "45 - 50",
    length_of_hire_rounded <= "55M 0S" & 
      length_of_hire_rounded > "50M 0S" ~ "50 - 55",
    length_of_hire_rounded <= "60M 0S" & 
      length_of_hire_rounded > "55M 0S" ~ "55 - 60",
    length_of_hire_rounded <= "3H 0S" & 
      length_of_hire_rounded > "1H 0S" ~ "1Hrs - 3Hrs",
    length_of_hire_rounded > "3H 0S" ~ "3Hrs plus",
  ))


durations <- a %>% 
  group_by(duration_block) %>% 
  summarise(count = n()) 
durations

time_blocks <- c("0 - 5", "5 - 10", "10 - 15", "15 - 20", "20 - 25", "25 - 30",
                 "30 - 35", "35 - 40", "40 - 45", "45 - 50", "50 - 55", 
                 "55 - 60", "1Hrs - 3Hrs", "3Hrs plus")

durations %>%   
  ggplot()+
  aes(x = duration_block, y = count) +
  scale_x_discrete(limits = time_blocks) +
  geom_col()+
  theme(axis.text.x = element_text(angle=45, hjust=0.9))

```



### GEOGRAPHICAL SPREAD 

Duration vs starting location 
There are 52 start stations and 55 end stations 
```{r}
start_coords <- nyc_bikes_df %>%
  unite("start_coords", start_lat, start_long, sep = ", ", remove = FALSE)
start_coords
```


```{r}
library(sf) # "simple features"
library(rgeos)
library(rnaturalearth)
library(rnaturalearthdata)
library(leaflet)
```


How many bikes from each station (start)
```{r}

# how many bikes from each station (52 start stations): 
bikes_per_start_station <- nyc_bikes_df %>% 
  group_by(start_long, start_lat, start_station) %>% 
  summarise(total_hires = n(), average_duration = round(mean(hire_duration))) %>% 
  arrange(desc(total_hires))
bikes_per_start_station


# 55 end stations
bikes_per_end_station<- nyc_bikes_df %>%
  group_by(end_long, end_lat, end_station) %>% 
  summarise(total_returns = n(), average_duration = round(mean(hire_duration))) %>% 
  arrange(desc(total_returns))
bikes_per_end_station

```




```{r}
#plot on scatter??
bikes_per_start_station %>% 
  ggplot()+
  aes(x = start_station, y = total_hires)+
  geom_point()

#there are too many, filter for most popular
bikes_per_start_station %>% 
  filter(total_hires > 90) %>% 
  ggplot()+
  aes(x = start_station, y = total_hires)+
  geom_col()
```





looking at radius:
```{r}
bikes_per_start_station %>% 
  leaflet() %>% 
  addTiles() %>% 
  addCircleMarkers(lng = ~start_long, 
                   lat = ~start_lat,
                   radius = ~total_hires/10,
                   popup = ~paste0("Start Station: ", start_station, 
                                   "<br> Total Hires: ", total_hires, 
                                   "<br> Average Duration: ", average_duration, 
                                   " mins"))
```


radius in clusters (though I think it's better to see out in the open)
```{r}
bikes_per_start_station %>% 
  leaflet() %>% 
  addTiles() %>% 
  addCircleMarkers(lng = ~start_long, 
                   lat = ~start_lat,
                   radius = ~total_hires/2,
                   clusterOptions = markerClusterOptions(),
                   popup = ~paste0("Start Station: ", start_station, 
                                   "<br> Total Hires: ", total_hires))
```


```{r}
bikes_per_end_station%>% 
  leaflet() %>% 
  addTiles() %>% 
  addCircleMarkers(lng = ~end_long, 
                   lat = ~end_lat,
                   radius = ~total_returns/10,
                   popup = ~paste0("End Station: ", end_station,
                                   "<br>Total Returns ", total_returns, 
                                   "<br> Average Duration: ", average_duration, 
                                   " mins"))
```



DURATION BY STATION 
```{r}
nyc_bikes_df %>% 
  group_by(start_station, type) %>%
  summarise(count = n())
  

```


